Algebraic Groups\vspace{-1ex}

### 2019-11-18 Algebraic Geometry/Algebraic-Geometry.tex
update

>>>>

\section{November 15th, 2019}
Thinking about quadradic forms (avoiding characteristic 2 for a moment): Any quadratif form $Q(x_0,\dots,x_n)$ is equivalent
(via a linear change of coordinates) to $x_0^2+\cdots+x_m^2$ (when $k=\bar k$) for some $0\le m\le n$. Thus $Q$ is nondegenerate 
if and only if $m=n$.

Think about the rational map $\pi:\P n\dashrightarrow \P m$ sending $(x_0:\cdots:x_n)\mapsto (x_0:\cdots:x_m)$. Notice this is not 
defined at $x_0=\cdots=x_m=0$, a linear subspace $L\subseteq \P n$ called the \textit{center of $\pi$}.
For instance, the map $\P3\dashrightarrow\P2$ is not defined at $(0:0:0:1)$. In affine coordinates $\A3\subseteq \P3$, 
where we set $x_3=1$. Here we think of setting $z=1$ (since we want to send $(x,y,z)\mapsto(x/z,y/z)=(x:y:z)$).

Looking at where points go in the plane $\A2\subseteq \A3\subseteq \P3$, we see that it is somehow mapped via the line through the origin. 
This is a ``pinhole camera'' projection of $\P3$ onto $\P2$.
\begin{prob}
	Given $\A3\to\A2$ via the map $(x,y,z)\mapsto (x,y)$, can this be recognized as the restiction of a projective morphism?
\end{prob}

Notice that given $\pi:\P3\dashrightarrow\P2$, the center is a conic given by $x_0^2+x_1^2+x_2^2=0$. This is a cone! Sweet!

Check this out: If we have the affine map $\A3\to\A1$ with the quadratic form $(x/y)^2+1=0$, consider the pullback 
under this map: the solutions to this in $\bbC$ are $x=\pm iy$, and for each of these points $\alpha$, the point $(x/y)-\alpha=0$ 
pulls back to $x-\alpha y=0$, a plane! But we get two roots, so two planes in $\P3\supset \A3$. What is the intersection of these? The center!

\subsection{Cremona}
We have the Cremona map $\P2\dashrightarrow\P2$ sending $(x,y,z)\mapsto(1/x,1/y,1/z)$. This is ``obviously'' a birational involution. 
At least involutivity is apparent.

But we can scale by $xyz$ and get the map $(x,y,z)\mapsto(yz,xz,xy)$. Notice it is undefined at $(1,0,0),(0,1,0)$, and $(0,0,1)$.
Now notice that the original map wasn't defined where \textit{any of the coordinates were zero} while the latter only requires that no more 
than two are zero. Why is this kosher?!

The idea here is that each of the lines $x=0$, etc are sent to POINTS $(1,0,0)$, which are then not in the domain of the cremona map.
Thus the idea of involution is no longer meaningful!

\begin{prob}
	Consider the form $Q=xy-zw$ and its zero set in $\P3$. Then look at its image under the map $(x,y,z,w)\mapsto(y,z,w)$. SHow $\pi$ is birational and write down an 
	inverse then compute $\pi\circ \pi^{-1}$ using cremona.
\end{prob}

\section{November 18th, 2019}
Apparently Hitler loved cake.

Today we are going to do some more examples. However, first we will consider 
\subsection{Last week's problems}
Consider the standard projection $\A3\to\A2$. We wanted to think of it as a linear map on projective spaces. Recall 
\begin{defn}
	A \textbf{rational map} $f:X\dashrightarrow Y$ is an equivalence of morphisms $f_U:U\to Y$ where $\varnothing\ne U\subseteq X$ is open. The equivalence 
	relation is that $f_U=f_V$ if $f_U|_{U\cap V}=f_V|_{U\cap V}$.
\end{defn}
What does it mean for a map to be birational? One idea could be
\begin{prob}
	Make a category $\mathbf{Bir}$ of irreducbile varieties with mapes being rational maps (eq. classes). Try to think about inverses in this category.
\end{prob}
\begin{defn}
	A rational map $f:X\dashrightarrow Y$ is \textbf{birational} if there exists a $g:Y\dashrightarrow X$ such that 
	\[fg=\id_Y\qquad gf =\id_X\]

	$f$ is \textbf{dominant} if there exists an open $V\subseteq Y$ that is contained in $\Im \tilde f$ for some representative $\tilde f$ of $f$.
\end{defn}
\begin{prob}
	Show that if $f:X\dashrightarrow Y$ is birational then there exist nonempty opens $U$ and $V$ in $X$ and $Y$, respectively such that $f$ induces an isomorphism $U\xrightarrow{\sim} V$ of varieties.
\end{prob}
\begin{defn}
	An irreducible variety is \textbf{rational} if it is birational to $\P n$ for some $n$.
\end{defn}
\begin{rmk}
	Roughly speaking, we think that ``most of $X$ can be uniquely described by $n$ algebraic parameters.'' Try to look at some examples! The points where this doesn't make sense is jsut the 
\end{rmk}
\begin{prob}
	I am repeating some of the old problems to refresh their importance: try figuring out how a quadric projects from $\P2$ to $\P1$ by the usual projection map. There is actually an isomorphism!
\end{prob}
We've been sniffing around the idea that any non-degenerate quadric $Q\subseteq \P n$ is rational. What about cublics? WHich are rational? The cubic curves (defined by cubic equations in $\P2$)
are genus one and NOT rational. The cubic surfaces (those in $\P3$) are apparently.

### 2019-11-18 Algebraic Groups/Algebraic-Groups.tex
update

\begin{document}
%make the title page
\title{Algebraic Groups\vspace{-1ex}}
\author{A course by Jarod Alper and Julia Pevtsova\\
Notes by Nico Courts}
\date{Autumn 2019/ Winter and Spring 2020}
\maketitle

@@

\begin{ex}
	If $G=\Gm$, then every irreducible representation is one-dimensional: $\Gm\to\Gm$ sends $t\mapsto t^\alpha$.
\end{ex}
A natural question one may ask: what can we say in general?
\subsection{The regular representation}
The regular representation is $\Gamma(G)$ with coaction given by comultiplicaton.
\begin{prop}
	And finite dimentional representation $V$ of $G$ embeds $V\subseteq \Gamma(G)^{\oplus n}$.
\end{prop}
\begin{prf}
	The coaction gives us a map 
	\[V\xrightarrow{\sigma}\Gamma(G)\otimes V=\Gamma(G)\otimes_k V_{\text{vs}}\cong \Gamma(G)^{\oplus\dim V}\]
	where $V_{vs}$ is the underlying vector space of $V$. Then the claim is that the overall map is injective and that it is a map of $G$-reps.

@@

>>>>

\begin{document}
%make the title page
\title{Algebraic Groups\vspace{-1ex}}
\author{A course by Jarod Alper and Julia Pevtsova\\
Notes by Nico Courts}
\date{Autumn 2019/ Winter 2020}
\maketitle

@@

\begin{ex}
	If $G=\Gm$, then every irreducible representation is one-dimensional: $\Gm\to\Gm$ sends $t\mapsto t^\alpha$.
\end{ex}
A natural question one may ask: what can we say in general?
\subsection{The regular representation}
The regular representation is $\Gamma(G)$ with coaction given by comultiplicaton.
\begin{prop}
	And finite dimensional representation $V$ of $G$ embeds $V\subseteq \Gamma(G)^{\oplus n}$.
\end{prop}
\begin{prf}
	The coaction gives us a map 
	\[V\xrightarrow{\sigma}\Gamma(G)\otimes V=\Gamma(G)\otimes_k V_{\text{vs}}\cong \Gamma(G)^{\oplus\dim V}\]
	where $V_{vs}$ is the underlying vector space of $V$. Then the claim is that the overall map is injective and that it is a map of $G$-reps.

@@

\section{November 13, 2019}
The big theorem here:
\begin{thm}
	Let $G$ be an algebraic group over $k$. Let $H\subseteq G$ be an algebraic subgroup (over $k$). Then the
	quotient $G/H$ exists as a quasi-projective scheme over $k$.

	Moreover if $H\subseteq G$ is normal, then $G/H$ is also an algebraic group.
\end{thm}

Today we are going to prove/see the case when $G$ is a linear algebraic group and smooth (reduced) and that $H\subseteq G$ is normal and $k=\bar k$.
The closedness of $k$ and the smoothness can be dropped at the expense of a couple extra lectures, probably.

\begin{defn}
	A map $G\xrightarrow{\pi} Q$ of algebraic groups over $k$ is a \textbf{quotient} if $\pi$ is faithfully flat.
\end{defn}

Using descent, we showed 
\begin{prop}
	If $G\xrightarrow{\pi} Q$ is a quotient, then  if we have another map $G\xrightarrow{\phi} P$ and furthermore that $\ker\pi\subseteq \ker\phi$,
	then we get a (unique) factorization $Q\to P$.
\end{prop}

We also have
\begin{prop}
	Any map of linear algebrai groups $G\xrightarrow{\phi} H$ factors as $G\to Q\hookrightarrow H$ where $G\to Q$ is the quotient map.
\end{prop}
\begin{prf}
	In the case that $G$ is reduced, we get a surjection onto $\Im\phi^\ast$ of $\Gamma(H)$, and $Q=\Spec\Im(\phi^\ast)$.
	Then it just takes showing that the surjection $\Gamma(H)\twoheadrightarrow \Im(\phi^\ast)$ induces a closed immersion of $Q$ in $H$.

	We also need a lemma that I missed but it can be found both in Milne and Waterhouse. This is where the heavy lifting is done.
\end{prf}
\section{November 15th, 2019}
Today we are going to be talking about properties of algebraic groups as well as properties of lements of $g\in G$. Today, we will be focusing on 
$g\in \GL(V)$, so essentially talking linear algebra.

\begin{defn}
	Let $V$ be a finite dimensional vector space over any field $k$. Let $g\in \GL(V)(k)$. Then $g$ is
	\begin{itemize}
		\item \textbf{diagonalizable} if there exists a basis of eigenvectors.
		\item \textbf{semisimple} if there exsits an extension $k'/k$ such that $g\otimes\id\in\GL(V\otimes_kk')$ is diagonalizable.
		\item \textbf{unipotent} if $g-\id$ is nilpotent.
		\item \textbf{triagonalizable} if there exists a basis so that $g$ is upper triangular.
	\end{itemize}
\end{defn}

Let $P_g\in k[T]$ be the characteristic polynomial for $g$. We say the \textit{eigenvalues of $g$ are in $k$} if $P_g$ splits over $k$. Thus if $g$ is diagonalizable, 
the eigenvalues are in $k$ and this is exactly equivalent to $g$ being triagonalizable. Furthermore $g$ is unipotent if and only if 
all of its eigenvalues are 1 or 0 (in $k$).
\begin{prop}
	If the eigenvalues $\lambda_i$ are in $k$, then 
	\[V\cong \oplus\ker(g-\lambda_i\id)^{a_i})\]]
	which gives us our Jordan decomposition in terms of generalized eigenspaces.
\end{prop}

\subsection{Jordan Decomposition}
\begin{thm}
	Let $k$ be perfect and let $g\in\GL(V)$ with $V$ finite dimensional over $k$. Then there exist unique $g_s$ and $g_u$ in $\GL(V)$ such that 
	\begin{itemize}
		\item $g=g_sg_u=g_ug_s$
		\item $g_s$ is semisimple and $g_u$ is unipotent.
	\end{itemize}
	Moreover, $g_s$ and $g_u$ are polynomials in $g$.
\end{thm}
\begin{prf}
	Assume first that $P_g(T)$ splits over $k$. Thus we can decompose 
	\[V\cong V_{\lambda_i}\]
	where $V_{\lambda_i}=\ker((g-\lambda_i\id)^{a_i})$.
	This gives us a Jordan decomposition of $g$ as a matrix with eigenvalues on the diagonal and is upper triangular. Take $g_s$ to be $\diag(\lambda_1,\cdots,\lambda_1,\dots,\lambda_s)$, 
	the diagonal of the matrix in this form. This is invertible (since $g\in\GL(V)$) so we set $g_u=g_s^{-1}g$. It is clear from this that these matrices commute and that they are 
	semisimple and unipotent, respectively.

	To get that these are polynomials in $g$, you go through an argument I missed. The uniqueness comes from looking at 
	\[g=g_sg_u=h_sh_u\]
	and taking 
	\[h_s^{-1}g_s=h_ug_u^{-1}\]
	and since everything commutes (check this), you get that these two elements are both semisimple and unipotent. But then they are diagonalizable with all 
	eigenvalues 1, so they are the identity and we are done.

	For the more general case, choose a finite extension $k'/k$ such that $P_g$ splits. Since $k$ is perfect, this is seperable. So we can assume it is Galois and 
	set $G=\Gal(k'/k)$. So $g\otimes \id\in\GL(V\otimes_k k')$ factors uniquely as $g\otimes \id=g_ug_s$ with $g_u,g_s\in\GL(V)(k')$.

	But then using the action of $G$ on $\GL(V)(k')$, by uniqueness of this decomposition $\sigma g_s=g_s$ and $\sigma g_u=g_u$ for all $\sigma\in G$ so 
	the elements are in fact in the field fixed by $G$ (i.e. $k$). Similarly $G$ acts on $Q(T)\in k'[T]$ but must fix the polynomials describing the factors as polynomials in $g$.
\end{prf}

As an example, consider a field $k$ of characteristic 2 and pick $\alpha\in k/k^2$ (obviously $k$ is not perfect). Then the matrix $g=(\begin{smallmatrix}
	0& 1\\\alpha,0
\end{smallmatrix})$ has polynomial $T^2-\alpha$, so with a single repeated eigenvalue $\sqrt a\notin k$.

\subsection{A broader context}
So then if $\iota:G\hookrightarrow\GL(V)$ is inclusion (faithful representation) we can decompose $\iota(g)=g_ug_s$ using the above results. Some properties of this phenomenon:
\begin{prop}
	If $L:V\to W$ is a linear transformation and $g\in\GL(V)$ and $h\in\GL(W)$ such that $L\circ g=h\circ L$,
	then $L\circ g_\ast=h_\ast\circ L$ for $\ast=u,s$.
\end{prop}
\begin{prop}
	\begin{align*}
		(g\oplus h)_\ast&=g_\ast\oplus h_\ast\\
		(g\otimes h)_\ast&=g_\ast\otimes h_\ast
	\end{align*}
	for $\ast=s,u$.
\end{prop}
\begin{prop}
	If $\rho:\GL(V)\to \GL(W)$ is a representation, then $\rho(g)_\ast=\rho(g_\ast)$.
\end{prop}

\begin{defn}
	For $g\in G(k)$, $g$ is \textbf{semisimple (resp. unipotent)} if for all representations $\rho:G\to \GL(V)$,
	$\rho(g)$ is semisimple (resp. unipotent).
\end{defn}

\section{November 18th, 2019}
\subsection{Generalizing the Jordan decomposition}
The further generalization of the Jordan decomposition we saw last time is
\begin{thm}
	If $k$ is perfect and $G$ is a linear algebraic group over $k$, then for any $g\in G(k)$, there exist unique $g_s,g_u\in G(k)$ such that 
	\begin{itemize}
		\item $g=g_sg_u=g_ug_s$
		\item $g_s$ is semisimple and $g_u$ is unipotent.
		\item $g_s$ and $g_u$ are polynomials in $g$.
	\end{itemize}
\end{thm}
Last time we proved this for $G=\GL(V)$. Then we saw a collection of propositions that gave us that semisimplicity and unipotence act nicely 
with respect to direct sum, tensor, and representations. This allowed us to define what semisimple and unipotent objects are in $G(k)$. This gave us 
a result which I think I botched last time:
\begin{prop}
	If $G$ is alinear algebraic group and $V$ is a faithful finite dimentional representation ($G\hookrightarrow \GL(V)$ is a closed embedding), then eery representation 
	of $G$ is obtained from $V$ using sums, tensors, subrepresentations, and quotients.
\end{prop}
\begin{rmk}
	An aside: what does it mean for a subspace $W\subset V$ to be fixed by $g\in G$ ($V\in \lmod G$)? If $V$ is finite-dimensional, then one can just take the image of $G$ in $\GL(V)$
	and ask whether $gW\subseteq W$.

	More generally since $g$ is a $k$-point, there is a maximal ideal $\frakm_g\lhd \Gamma(G)$ such that 
	\[\Gamma(G)/\frakm_g\cong k\] 
	with quotient map $q$. Then this gives us a map $\Gamma(G)\otimes V\xrightarrow{q\otimes\id} k\otimes V\cong V$ which, when composed with the coaction $V\to\Gamma(G)\otimes V$,
	gives us a map $m_g:V\to V$. Then we can say that $W$ is fixed by $g$ if $m_g(W)\subseteq W$.
\end{rmk}
\begin{prf}[in general]
	Since $G$ is linear algebraic, we can embed $G\hookrightarrow\GL_n$ for some $n$. This gives us a way to decompose $g=g_sg_u$, but the question is 
	whether $g_s$ and $g_u$ actually lie in (the image of) $G$. Thus for every representation $\rho:G\to \GL(W)$, we get a pair of elements $\rho(g)_s$ and $\rho(g)_u$. Due to the properties we saw before, 
	these are nice.

	The idea here is that one representation we have at our disposal is the regular representation. In general this is large, but it recovers our group $G$ in some way and this will be key to our discussion.
	Let $V=\Gamma(\GL_n)$, the regular representation of $\GL_n$. Then we get maps 
	\[G\hookrightarrow\GL_n\to\GL(V)\]
	and we have an ideal $I\subseteq \Gamma(GL_n)$ where $I$ is the ideal defining $G$. Then the claim is that $g\in G(k)$ stabilizes $I$. This is in Waterhouse chapter 9.
\end{prf}

\subsection{Group properties}
This enables us to make more classifications of groups:
\begin{prop}
	We say that $G$ is \textbf{diagonalizable} if there is a faithful representation $G\hookrightarrow\GL_n$ that factors through the 
	diagonal matrices in $\GL_n$.
\end{prop}
\begin{rmk}
	This is equivalent to the earlier definition of there being a closed embedding of $G\hookrightarrow\Gm^n$.
\end{rmk}
\begin{defn}
	We say that $G$ is of \textbf{multiplicative type} if $G\times_k\bar k$ is diagonalizable.
\end{defn}

\begin{prop}
	If $G$ is a commutative linear algebraic group over $K$ and all elements $g\in G(k')$ for $k\to k'$ a field extension are semisimple, then $G$ is of multiplicative type.
\end{prop}
\begin{prf}
	We can assume that $k=\bar k$ and we know that all $g\in G(k)$ are diagonalizable. Then we use the fact that commuting diagonalizable 
	matrices are simultaneously diagonalizable. The result follows.
\end{prf}
\begin{rmk}
	The proof of the linear algebra fact above comes from taking your commuting diagonalizable elements $g$ and $h$ and showing that each 
	$\lambda$ eigenspace of $g$ is fixed by $h$ and vice versa, giving us the same number and dimension of eigenspaces.
\end{rmk}
